---
title: "Ageprediction"
author: "Kaisu Hiltunen"
date: "2024-01-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Age prediction
Age prediction was done with scEpiAge blood clock.
Used parameters
tissue="blood"
nSimulations = 5
```{r, warning=FALSE, message=FALSE}
setwd("C:/Users/khiltunen/Documents/R/scEpiAge")
options(bitmapType='cairo')
library(ggplot2)
library(dplyr)
library(VennDiagram)
```


```{r}
#colorblind colors '#CC6677', '#332288', '#DDCC77', '#117733', '#88CCEE', '#882255', '#44AA99', '#999933', '#AA4499'. Bad data: '#DDDDDD'

#colors ("#2978a0", "#710C32", "#fc8d62", "#68CBAC")
tissue = "blood"
file <- "extended_predS183_blood.txt"
```


```{r functions for age predictions}
create_predictions_df <- function(file){
  # Function for reading the age prediction data into a dataframe. 
  data <- read.delim(file)
  
  #extract tissue txpe
  tissueType <- sub(".*_(\\w+)_r\\d+_.*", "\\1", rownames(data))
  data$TissueType <-tissueType
  
  return(data)
}
# Age deviation is actually the same as absolute error
calculate_avg_errors <- function(data){
  # Function to calculate average errors by tissue type 
  avg_error_by_tissue <- data %>%
    group_by(TissueType) %>%
    summarize(AvgAbsoluteError = mean(ageDeviation))
  return(avg_error_by_tissue)
}

calculate_avg_relative_errors <- function(data){
  # Function to calculate average relative errors by tissue type 
  # This function takes into count how large the error is compared to the 
  # actual age
  
  #Calculate relative error
  data$relativeError <- 
    ((data$predictedAge - data$actualAge)/data$actualAge)*100
  
  avg_r_error_by_tissue <- data %>%
    group_by(TissueType) %>%
    summarize(AvgRelativeError = mean(relativeError))
  
  return(avg_r_error_by_tissue)
}

residual_hist <- function(data){
  #Function to plot the frequency histogram of residuals i.e. the agedeviations 
  
}

age_pred_plot <- function(data, clock){
  # Plot actual age vs. predicted age using ggplot
  ggplot(data, aes(x = actualAge, y = predictedAge)) +
    geom_point(aes(color = TissueType, shape = TissueType), 
               alpha=0.7, size=3) +
    scale_color_manual(values = c("BAT" = "#710C32", 
                                  "Blood" = "#2978a0", 
                                  "Liver" = "#68CBAC", 
                                  "scAT" = "#fc8d62")) +
    scale_shape_manual(values = c("BAT" = 19, "Blood" = 17, 
                                  "Liver" = 15, "scAT" = 8)) +
    # Add linear trendlines
    geom_smooth(aes(color=TissueType),
                method = "lm", se = FALSE, 
                size = 0.7, linetype="dashed") +  
    labs(title = paste0("Age prediction (", clock, " clock)"), 
         x="Chronological age", 
         y= "Predicted age", 
         color="Tissue regression") +
    geom_abline(intercept = 0, slope = 1, linetype = "solid", color = "black")+
    guides(color = guide_legend(override.aes = list(shape = c(19, 17, 15, 8)),
                                title = "Tissue sample",
                                title.position = "top"), shape = FALSE)+
    theme_minimal()+
    theme(legend.background = element_rect(fill = "white", colour = "grey", linewidth = 0.5),
          legend.justification = c(0, 1),  # Justify legend position to the top-right
          legend.position = c(0, 1),       # Position legend inside the plot at the top-right
          )
}

# Function to plot ageDeviation vs. actual age using ggplot
deviation_plot <- function(data, clock){
  if (length(unique(data$TissueType)) == 4){
    shapelist = c(19, 17, 15, 8)
  }
  else if (length(unique(data$TissueType)) < 4){
    shapelist = c(17)
  }
  ggplot(data, aes(x = actualAge, y = ageDeviation, 
                   color = TissueType,shape = TissueType)) +
    geom_point() +
    labs(title = paste("Absolute error:",clock,"clock"),
           x="Chronological age", 
           y= "Absolute error", 
           color="Tissue" ) +
    scale_shape_manual(values = c("BAT" = 19, "Blood" = 17, 
                                  "Liver" = 15, "scAT" = 8)) +
    scale_color_manual(values = c("BAT" = "#710C32", 
                                    "Blood" = "#2978a0", 
                                    "Liver" = "#68CBAC", 
                                    "scAT" = "#fc8d62"))+
      # Add linear trendlines
    geom_smooth(aes(color=TissueType),
                  method = "lm", se = FALSE, size = 0.7, linetype="dashed") +
    guides(shape = FALSE)+
    guides(color = guide_legend(override.aes = list(shape = shapelist),
                                       title = "Tissue sample",
                                       title.position = "top"), 
                                       shape = FALSE)+
    theme_minimal()
}

#Function for getting the rownames to make a Venn diagram of the clock sites
get_rownames <- function(tissue){
  if(tissue=="liver"){
  expectedMethMatrix <- 
    read.delim("./ExpectedMethylationMatrices/ExpectedMethMat_Liver.tsv",
               as.is=T,row.names=1,check.names = F)
  } else if(tissue=="lung"){
  expectedMethMatrix <- 
    read.delim("./ExpectedMethylationMatrices/ExpectedMethMat_Lung.tsv",
               as.is=T,row.names=1,check.names = F)
 } else if(tissue=="blood"){
  expectedMethMatrix <- 
    read.delim("./ExpectedMethylationMatrices/ExpectedMethMat_Blood.tsv",
               as.is=T,row.names=1,check.names = F)
 }
  return(rownames(expectedMethMatrix))
}

```



**Visualization:**
Here linear regression plots are produced to provide visualizations of the relationships between actualAge and predictedAge. 

Make plot to compare frequency of pred age - actal age between tissues.
```{r blood clock data}
# Retrieve predictions
blood_clock_data <- create_predictions_df("extended_predS183_blood.txt")
# Calculate errors
errors <- calculate_avg_errors(blood_clock_data)
relative_errors <- calculate_avg_relative_errors(blood_clock_data)

age_pred_plot(blood_clock_data, "Blood")
```


# Liver clock predictions
```{r liver clock data}
# Retrieve predictions
liver_data <- create_predictions_df("extended_predS183_liver.txt")
# Calculate errors
errors <- calculate_avg_errors(liver_data)
relative_errors <- calculate_avg_relative_errors(liver_data)

age_pred_plot(liver_data, "liver")


```


```{r lung clock data}
# Retrieve predictions
lung_data <- create_predictions_df("extended_predS183_lung.txt")
# Calculate errors
errors <- calculate_avg_errors(lung_data)
relative_errors <- calculate_avg_relative_errors(lung_data)

age_pred_plot(lung_data, "lung")
```






```{r ageDeviation blood clock data}
# Make a subset for blood sample age predictions with blood clock
subset_blood_data <- blood_clock_data %>%
  subset(TissueType == "Blood")
# Plot age deviation for this subset
deviation_plot(subset_blood_data, "Blood")
```
```{r abs error histogram}
deviation_histogram <- function(data, clock) {
  ggplot(data, aes(x = predictedAge-actualAge, fill = TissueType)) +
    geom_histogram(binwidth = 6, position = "dodge", alpha = 0.7) +
    labs(title = paste("Error histogram:", clock, "clock"),
         x = "Error",
         y = "Frequency",
         fill = "Tissue") +
    scale_fill_manual(values = c("BAT" = "#710C32",
                                 "Blood" = "#2978a0",
                                 "Liver" = "#68CBAC",
                                 "scAT" = "#fc8d62")) +
    theme_minimal()
}

# Example usage:
deviation_histogram(blood_clock_data, "Blood")

```

```{r ageDeviation lung clock data}
deviation_plot(blood_clock_data, "Blood")
```

```{r ageDeviation lung clock data}
deviation_plot(lung_data, "lung")
```
```{r ageDeviation liver clock data}
deviation_plot(liver_data, "liver")
```
#IQR analysis

```{r}
IQR_ratio_histogram <- function(data, clock) {
  ggplot(data, aes(x = IQR_Ratio, fill = TissueType)) +
    geom_histogram(binwidth = 4, position = "dodge", alpha = 0.7) +
    labs(title = paste("IQR histogram:", clock, "clock"),
         x = "IQR Ratio",
         y = "Frequency",
         fill = "Tissue") +
    scale_fill_manual(values = c("BAT" = "#710C32",
                                 "Blood" = "#2978a0",
                                 "Liver" = "#68CBAC",
                                 "scAT" = "#fc8d62")) +
    theme_minimal()
}
IQR_ratio_histogram(blood_clock_data, "Blood")


```


# R-squared analysis


```{r}
# Create a function to calculate R-squared for a specific tissue group
calculate_r_squared <- function(data) {
  # Step 1 Calculate the Mean Squared Error (MSE) of your model
  mse_model <- mean((data$actualAge - data$predictedAge)^2)

  # Step 2 Calculate the MSE of the mean (baseline model)
  mse_mean <- mean((data$actualAge - mean(data$actualAge))^2)

  # Step 3 Calculate R-squared
  r_squared <- 1 - (mse_model / mse_mean)
  
  return(r_squared)
}

# Initialize an empty list to store results
r_squared_results_list <- list()

# Get unique tissue types
unique_tissues <- unique(blood_clock_data$TissueType)

# Iterate over each tissue group
for (tissue in unique_tissues) {
  # Filter data for the current tissue
  current_data <- dplyr::filter(blood_clock_data, TissueType == tissue)
  
  # Calculate R-squared for the current tissue
  r_squared_value <- calculate_r_squared(current_data)
  
  # Store the result in the list
  r_squared_results_list[[tissue]] <- r_squared_value
}

# Convert the list to a data frame
r_squared_results <- data.frame(
  R_squared = unlist(r_squared_results_list)
)

# Print the summarized results
print(r_squared_results)

```






# Clock sites analysis

```{r Clock sites}
# compare the number of sites
# Read data files
data_blood <- read.delim("extended_predS183_blood.txt")
data_liver <- read.delim("extended_predS183_liver.txt")
data_lung <- read.delim("extended_predS183_lung.txt")
data_lung$clockType <- "Lung"
data_blood$clockType <- "Blood"
data_liver$clockType <- "Liver"
#print(data_lung)
# Combine the data frames
data_combined <- rbind(data_blood, data_liver, data_lung)
# Plot

# Organize clockType in the desired order
data_combined$clockType <- factor(data_combined$clockType, levels = c("Blood", "Lung", "Liver"))

# Plot a boxplot
site_plot <- ggplot(data_combined) +
  geom_boxplot(aes(x = clockType, y = sitesUsed, fill = clockType)) +
  labs(title = "Number of Sites Used in Predicition", fill="Clock", y="Sites Used by Clock", x="Clock type") +
  scale_fill_manual(values = c("Blood" = "#2978a0", 
                                "Liver" = "#68CBAC", 
                                "Lung" = '#88CCEE'))+
  theme_minimal()
  
site_plot
```
```{r clock sites overlap}
blood_sites <- get_rownames("blood")
liver_sites <- get_rownames("liver")
lung_sites <- get_rownames("lung")

# Create a list of sets
venn_list <- list(
  "blood" = blood_sites,
  "liver" = liver_sites,
  "lung" = lung_sites
)

# Create the Venn diagram
venn <- venn.diagram(venn_list,filename = 'Euler_3set_simple.png')
venn
?venn.diagram
# Customize the plot if needed
grid.draw(venn.plot)

```


